 

### 2020-12-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming. As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [Bartholomew et al. 2020]. Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011]. Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[Rollinson et al. 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]. Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures. 
Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [McDowell et al. 2020].
Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al. 2020). Each of these changes have potential feedbacks to climate change.

@@

**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al. 1999]. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[Bauerle et al. 2011]. In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999]. Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; Stockes et al. 2006].

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [De Frenne et al. 2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].

@@

Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance with intensity and availability of light. Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al. 2018; @niinemetsWorldwideAnalysisWithincanopy2015]. This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [Sack et al. 2006; Ishii et al. 2018].

@@

Cateogorization of sun and shade tolerance from Niinemetes and Valladares, 2006: "The five-level scale used for shade tolerance (1,
very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%"

@@

Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

>>>>

While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming. As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [@bartholomewSmallTropicalForest2020]. Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011]. Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[Duque et al. 2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]. Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures. 
Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

@@

**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999]. Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006].

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].

@@

Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance with intensity and availability of light. Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al. 2018; @niinemetsWorldwideAnalysisWithincanopy2015]. This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].

@@

Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006: "The five-level scale used for shade tolerance (1,
very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%"

@@

Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [@chenLeafEconomicsSpectrum2020a], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
